### 2015-03-27 markdown/manuscript.tex
second pass, part 2

\section{How we treat GP (and how it treats us in turn)}\hypertarget{how-we-treat-gp-and-how-it-treats-us-in-turn}{}\label{how-we-treat-gp-and-how-it-treats-us-in-turn}

Genetic Programming (and the broader body of work to which it belongs, which I prefer to call ``generative processing'' and will also abbreviate as ``GP'') embodies a very particular \emph{stance} towards the scientific and engineering work of modeling, design, analysis and---though least of all---optimization. A case could be made that the resistance we have all recounted towards GP from the technical and lay audience has little to do with the technical results so amply demonstrated over the last quarter-century, but rather from discomfort among that audience in GP's particular ``way of working'' on problems. There is a tacit assumption, even among GP theorists and practitioners, that science and engineering are ``rigorous'', or even successful, only when they proceed through ordered phases of

\begin{enumerate}
\item conceptualization, the ``vision thing''
\item planning
\item design
\item architecture
\item implementation
\item testing
\item debugging
\end{enumerate}

Very few scientists or engineers would admit that any \emph{real} project has followed this narrative structure in practice, but in every phase of work life from fund-raising to published reports the story is framed in that way, especially in scientific work: Because of work gone before, an idea or insight was had. The idea was framed as a formal hypothesis. The hypothesis (and current Best Statistical Practices) suggested an experimental design obvious to any in the discipline. The experimental design was undertaken, the data collected, the hypothesis tested, and we can be confident of its veracity because, well, you just heard me say ``Best Statistical Practices'', right?

Under trivial term substitutions---``cost--benefit analysis'' and ``requirements document'' for ``hypotheses'' and ``experimental design'', for example---the same narrative can be used to describe any institutional project management and public policy planning process as well. The flow in every case is essentially from \emph{vision} to \emph{plan}, \emph{plan} to \emph{implementation}, \emph{implementation} to \emph{verification}, and \emph{verification} to \emph{validation}.

Of course, nobody ``really believes'' this narrative who has ever done any of the work. It is a matter for another day to draw parallels with the social construction of religious belief.\footnote{Paul Veyne's excellent \emph{Did the Greeks Believe in Their Myths?} might be an interesting starting point, though.} There have been numerous philosophical challenges to this artificial narrative of course, from Peirce and Dewey nearly a century ago, to Kuhn and Lakatos in the 1970s, and more today.

@@

Andrew Pickering's \emph{Mangle of Practice} is the one I'd like to use here. This approach is noteworthy because it is so close to our actual experience using GP, of the ``didn't we already know this?'' sort. At the risk of eliding a lot of extraordinarily well-considered structure, let me summarize:

The act of ``doing science'' is \emph{at no point whatsoever} a behavior undertaken by an isolated ``researcher'' in an objective field of externalities. Rather, the researcher who initiates this process with a vision, intuition, hypothesis or hunch begins always by \emph{making some artifact}: code, an instrument, a proof, a sketch, a maquette or even a thoughtful conversation at a conference.

The thing, inevitably, \emph{resists}. That is, Pickering grants it agency, or rather makes it the agent of all those externalities that impinge on the work to make the \emph{vision} differ from \emph{practice}: the facts of the actual world, the cultural assumptions and habits of discipline, the raw materials and toolkit available to the practitioner, and so forth. ``Resistance'' here is not merely the obvious trouble of a software bug or some missing parts, but includes one's sense \emph{on seeing it} that something's not quite right; the realization that more (or less) is needed. The language we use, in the face of this resistance, is that the \emph{thing made} ``feels wrong'' or ``points something out'', that it ``wants to do X instead of Y'', or that ``it's doing something too complicated for me to understand right now''.

And---assuming science is indeed what's being done---the researcher \emph{changes in response to this resistance}. That vision changes, the plan adapts, or in some other way the \emph{thing made} causes a response in the state of the researcher herself. Pickering's Mangle is this emergent dance of inanimate agency: the researcher starting to follow a vision by making (or altering) a \emph{thing}, and the \emph{thing made} in turn acting as a channel for the world itself to steer the researcher in another direction.

The ``mangling'' of Pickering's metaphoric name refers not to wounding but to the mechanical laundry apparatus, the antique wringer through which sheets and linens are run to drain the water, and which as a \emph{side-effect} impose a novel higher-dimensional structure and juxtapose unexpected components with one another. Here, the response of the researcher to resistance she experiences in the course of her project is no less a part of the dynamics of ``science'' than the act of writing and running code, the authority warranted by particular statistical practices in her discipline, or the raw pressure of physical laws.

@@

The broader field of machine learning leans seems to take a much more traditional stance towards its subject matter: machine learning frameworks (excepting GP) are each discrete tools aimed at producing standardized and reproducible results to particular statistical questions. The result of training a neural network or random forest on a given data set is not expected to be a \emph{surprise} in any real sense, but rather the sufficiently robust result of applying numerical optimization to a particular mathematical programming model. Whether one describes the process as minimizing out-of-sample error or maximizing information gain, the focus of the discipline is on contingent reliability rather than exploratory modeling.

>>>>

\section{How we treat GP, and how it treats us in turn}\hypertarget{how-we-treat-gp-and-how-it-treats-us-in-turn}{}\label{how-we-treat-gp-and-how-it-treats-us-in-turn}

Genetic Programming\footnote{And not just Genetic Programming as such, but also the broader discipline to which I claim it belongs and which is not obliged to be either ``genetic'' or ``programming''. I prefer to call this looser collection of practices ``generative processing'', and will also abbreviate it ``GP''; assume I mean the latter in every case.} embodies a very particular \emph{stance} towards the scientific and engineering work of modeling, design, analysis and optimization. I increasingly suspect the resistance we've all recounted towards GP from our prospective technical and lay audience has little to do with our technical results, but rather arises from uncertainty among that audience with GP's very particular ``way of working''.

There is a tacit assumption, even among GP theorists and practitioners, that science and engineering are only ``rigorous'' when they proceed through a sequence of ordered phases from planning to implementation. The ``scientific method'' is most often represented something along the lines of

\begin{quote}
\begin{enumerate}
\item conceptualization; 2. planning; 3. design; 4. architecture; 5. implementation; 6. testing; 7. debugging
\end{enumerate}
\end{quote}

I am sure very few scientists or engineers of my acquaintance would admit any \emph{real} project (in history) has ever followed this narrative arc in a literal sense. But that story nonetheless informs and constrains much of our work lives, from fund-raising to publishing reports:

\begin{quote}
Because of the body of published work, an insight was had. The insight was framed as a formal hypothesis. The hypothesis (and current Best Statistical Practices) suggested an experimental design, one familiar and obvious to any in Our Discipline. The experimental design was undertaken, the data were collected, the hypothesis duly tested, and now we can be confident of its veracity because\ldots{} well, you just heard me say ``Best Statistical Practices'', right?
\end{quote}

Under trivial term substitutions---``cost--benefit analysis'' and ``requirements document'' for ``hypotheses'' and ``experimental design'', for example---the same narrative can be used to describe almost any institutional project management and public policy planning process as well. The flow in every case is essentially from \emph{vision} to \emph{plan}, \emph{plan} to \emph{implementation}, \emph{implementation} to \emph{verification}, and \emph{verification} to \emph{validation}.

Of course, nobody ``really believes'' this narrative who has ever done any of the work. It is a matter for another day to draw parallels with the social construction of religious belief.\footnote{Paul Veyne's excellent \emph{Did the Greeks Believe in Their Myths?} might be an interesting starting point, though.} There have been numerous philosophical challenges to this artificial narrative of course, from Peirce and Dewey nearly a century ago, to Kuhn and Lakatos in the 1970s, and many more to be found in the Table of Contents in any Philosophy of Science text.

One in particular is my focus today: Andrew Pickering.

@@

Andrew Pickering's \emph{Mangle of Practice} is a decade old, but surprisingly little-known outside the field of Science Studies. His approach is especially noteworthy here because I find it so close to our actual experience using GP. Indeed, most colleagues who hear it for the first time utter an inevitable ``didn't we already know this?'' I think the distinction Pickering's approach is able to highlight is exactly the problematic one in our work: between the illusory (but publishable) narrative of ``scientific method'', and the realized experience we have of \emph{performing science}. At the cost of glossing a lot of his well-considered structure, let me summarize.

First, the act of ``doing science'' is \emph{at no point whatsoever} to be understood as one in which an isolated ``researcher'' works in an objective and static field of ``externalities'' and ``facts''. Rather, research begins only when the researcher undertakes to \emph{makes some artifact}: a block of code, a technical instrument, an equation, a sketch, a maquette or even a thoughtful conversation at a conference. Everything before the researcher begins to make these things is a matter of building ``vision'' (my term); research proper only occurs when work is done in the real world. I'll call this work product \emph{the thing made}, and keep in mind that it is a proxy for the vague collection we might otherwise call ``the project''.

We see Pickering's model moving quickly away from more traditional views of science when we realize he has given the inanimate \emph{thing made} an agency of its own. In any real project, we perceive the thing made \emph{resisting}---whether we imagine that it resists ``in itself'', or as an agent of externalities that impinge on the work in progress to make our driving vision differ from reality. The thing made comes to represent, in other words, the facts of the actual world, the cultural assumptions and norms of one's discipline, the raw materials and toolkit available to a practitioner, and so forth. ``Resistance'' here is not merely a problem with a software bug or lost raw materials, but includes one's sense \emph{on seeing it} that something's not quite right. The thing made's resistance, as perceived by the researcher, leads her to react in turn.

Consider the language we use in the face of this resistance: it ``feels wrong''; it ``points something out''; it ``wants to do X instead of Y''; it's ``doing something too complicated for me to understand right now''.

And---assuming science, engineering, art or any other creative process is indeed what's being done---the researcher \emph{changes in response to this resistance}. That vision changes, the plan adapts, or in some other way the \emph{thing made} causes a response in the state of the researcher herself. Pickering's Mangle is this emergent dance of inanimate agency: the researcher starting to follow a vision by making (or altering) a \emph{thing}, and the \emph{thing made} in turn acting as a channel for the world itself to steer the researcher in another direction.

Pickering's ``mangle''\footnote{The word ``mangle'' he has chosen is itself interesting and insightful:

\begin{quote}
\ldots{}I find ``mangle'' a convenient and suggestive shorthand for the dialectic because, for me, it conjures up the image of the unpredictable transformations worked upon whatever gets fed into the old-fashioned device of the same name used to squeeze the water out of the washing. It draws attention to the emergently intertwined delineation and reconfiguration of machinic captures and human intentions, practices, and so on. The word ``mangle'' can also be used appropriately in other ways, for instance as a verb. Thus I say that the contours of material and social agency are mangled in practice, meaning emergently transformed and delineated in the dialectic of resistance and accommodation\ldots{}.
\end{quote}} is this emergent dance of agency, undertaken between the researcher and the project: the researcher makes, the thing made resists, the researcher is influenced and redirected, \emph{accommodating} the resistance. It may seem glib to say that ``no plan survives contact with the enemy,'' but Pickering's narrative of creative work emphasizes the fact that no \emph{revised} plan survives unchanged, either. In the traditional narrative, we elide the work as it unfolded and re-frame it as a sort of idealized, apersonal Platonic truth: we use the passive voice, we hide the missteps and confusion, we paint a story moving from vision to plan to success.

In a GP setting, the many modes of resistance we perceive are the very ``symptoms'' Rick mused about. Even though we as researchers know we've written all the code and set all the parameters, we're nonetheless willing to speak of a GP run ``doing'' things, as opposed to merely unfolding according to our plan. A GP system does not ``resist'' by, for example, ``having the wrong population size''. Rather it resists by \emph{causing concern or dissatisfaction in the user}, which in turn sparks in that user a practical (if only explanatory) response, which leads them to change the \emph{thing made}.

@@

The broader field of machine learning seems to take a much more traditional stance towards its subject matter: machine learning frameworks (excepting GP) are each discrete tools aimed at producing standardized and reproducible results to particular statistical questions. The result of training a neural network or random forest on a given data set is not expected to be a \emph{surprise} in any real sense, but rather the sufficiently robust result of applying numerical optimization to a particular mathematical programming model. Whether one describes the process as minimizing out-of-sample error or maximizing information gain, the focus of the discipline is on contingent reliability rather than exploratory modeling.
